Well-quasi-orderings on word languages

### 2024-10-16 src/infixes.tex
Draft: proof of decidability for infix

\begin{theorem}
    \label{infix-embedding:thm}
    Let $(X, \preceq)$ be a \kl{partially ordered} set,
    and $\Sigma \defined \set{a,b,c}$.
    Then the following are equivalent:
    \begin{enumerate}
        \item \label{infix-embedding-embeds:item} 
            $X$ \kl{embeds into} $(\Sigma^*, \infleq)$,
        \item \label{infix-embedding-count:item}
            $X$ is countable, and for every $x \in X$,
            its \kl{downwards closure}
            $\dwset[\preceq]{x}$ is finite.
    \end{enumerate}
\end{theorem}
\begin{proof}
    Let us first prove that \cref{infix-embedding-embeds:item} implies
    \cref{infix-embedding-count:item}. Let $f \colon X \to \Sigma^*$ be an
    \kl{embedding}. Note that $f$ is injective, because $X$ and $\Sigma^*$
    are \kl{partially
    ordered}. In particular, because $\dwset[\infleq]{w}$ is finite
    for every $w \in \Sigma^*$, then so is $\dwset[\preceq]{x}$
    for every $x \in X$. Similarly, $X$ must be countable
    because $\Sigma^*$ is.

    Conversely, let us prove that \cref{infix-embedding-count:item} implies
    \cref{infix-embedding-embeds:item}. To that end, let us first define
    inductively a sequence $\seqof{X_n}$ of subsets of $X$ as follows: $X_0$ is
    the set of minimal elements of $X$, and $X_{n+1}$ is the set of minimal
    elements of $X \setminus \bigcup_{i \leq n} X_i$. Because initial segments
    of $X$ are finite, we know that $X = \bigcup_{n \in \Nat} X_n$. To
    simplify notations, we will also use $Y_n \defined \bigcup_{i \leq n} X_i$.
    We also let $L_0 \defined a b^* a$, and 
    $L_{n+1} \defined L_n \cup c^{n+1} b^* c^{n+1} (L_n c^{n+1})^*$.
    Now, let us inductively define \kl{embeddings} $f_n \colon (Y_n,\preceq) \to
    (L_n, \infleq)$.

    For the base case, let $\seqof{x_i}[i \in \Nat]$ be an enumeration of
    $X_0$, which is possible because $X$ is countable. We define $f_0 \colon
    X_0 \to \Sigma^*$ as follows: $f_0(x_i) \defined a b^i a$. It is immediate
    that $f_0$ is an \kl{embedding}. For the inductive step, let $\seqof{x_i}[i
    \in \Nat]$ be an enumeration of $X_{n+1}$, and assume that $f_n$ is an
    \kl{embedding} from $Y_n$ to $\Sigma^*$.
    Let us define $D_i \defined \setof{ y \in Y_n }{ y \preceq x_i }$
    for $i \in \Nat$, all of which are finite by assumption.
    We then let for all $i \in \Nat$:
    \begin{equation*}
        f_{n+1}(x_i) \defined 
        \underbrace{c^{n+1} b^i c^{n+1} \prod_{y \in D_i} \overbrace{f_n(y)}^{\in L_n} c^{n+1}}_{\in L_{n+1}}
        \quad .
    \end{equation*}
    And let $f_{n+1}(y) \defined f_n(y)$ for all $y \in Y_n$.

    To check that $f_{n+1}$ is an \kl{embedding}, let us first notice that
    words $f_{n+1}(x_i)$ and $f_{n+1}(x_j)$ are incomparable for the \kl{infix
    relation} when $i \neq j$, because the only factor of the form $c^{n+1}
    a^{n+1} b^t a^{n+1} c^{n+1}$ appears at the beginning of these words, and
    precisely encode $i$ (resp. $j$) in the value of $t$. Now, let us consider
    $y \in Y_n$ and $x_i \in X_{n+1}$ such that $y \preceq x_i$. Because $y \in
    D_i$, we have $f_n(y) \infleq f_{n+1}(x_i)$ by construction, and since
    $f_n(y) = f_{n+1}(y)$, we conclude that $f_{n+1}(y) \infleq f_{n+1}(x_i)$.
    Conversely, assume that $f_{n+1}(y) \infleq f_{n+1}(x_i)$. Again, this
    means that $f_n(y) \infleq f_{n+1}(x_i)$. Now, the encoding is robust
    enough so that we can conclude $f_{n}(y) \infleq f_{n}(z)$ for some $z \in
    D_i$ By induction hypothesis, $y \preceq z$, hence, $y \preceq x_i$.

    We conclude by noticing that $\bigcup_{n \in \Nat} f_n$ is an \kl{embedding}
    of $X$ into $\Sigma^*$.
\end{proof}

\begin{figure}
    \centering
    \includestandalone[width=\linewidth]{fig/infix-encoding-standalone}
    \caption{Representation of the \kl{subword relation} for $\set{a,b}^*$
        inside the \kl{infix relation} for $\set{a,b,c}^*$
        using the encoding of \cref{infix-embedding:thm}, restricted to words
        of length at most $3$. To obtain smaller words,
        we replaced $D_i$ by $\max D_i$ in this construction.
    }
    \label{infix-embedding:fig}
\end{figure}

The encoding using three distinct letters is clearer to use and to
illustrate, but it is not surprising that one can encode a three
letter alphabet into a two letter one, immediately
leading to the following \cref{infix-embedding-binary:cor}.

\begin{corollary}
    \label{infix-embedding-binary:cor}
    Let $(X, \preceq)$ be a \kl{partially ordered} set,
    and $\Sigma \defined \set{a,b}$.
    Then the following are equivalent:
    \begin{enumerate}
        \item \label{infix-embedding-embeds:item} 
            $X$ \kl{embeds into} $(\Sigma^*, \infleq)$,
        \item \label{infix-embedding-count:item}
            $X$ is countable, and for every $x \in X$,
            its \kl{downwards closure}
            $\dwset[\preceq]{x}$ is finite.
    \end{enumerate}
\end{corollary}
\begin{proof}
    Let us construct an \kl{order embedding} from 
    $(\set{a,b,c}^*, \infleq)$ to $(\set{0,1}, \infleq)$.
    To that end, consider
    the morphism $\mu$ that maps $a$ to $101$, $b$ to $1001$
    and $c$ to $10001$.
    Assume that $u \infleq v$ in $\set{a,b,c}^*$,
    then $\mu(u) \infleq \mu(v)$ because $\mu$ is a morphism.
    Conversely, assume that $\mu(u) \infleq \mu(v)$.
    That, is, there exists an \kl{infix} $w$ of $\mu(v)$
    such that $\mu(u) = w$.
    Because $\mu(u)$ starts with $10$ and ends with $01$, so does
    $w$. In particular, the word $w$ is compatible with $\mu$
    in the sense that $w = \mu(s)$ for some $s \infleq v$.
    Because the map $\mu$ is injective,
    we conclude that $\mu(s) = \mu(u)$ implies $s = u$
    and $u \infleq v$ as expected.
\end{proof}

@@

@@

The abstract definition of \kl{amalgamation system} can be enriched by
effectiveness assumptions, as explained in \cite[Section 3.1.4]{ASZZ24}. In
this setting, we conjecture that the following decidability result holds.

\begin{conjecture}
    Let $L$ be given by an \kl{effective amalgamation system},
    then it is decidable whether $L$ 
    is \kl{well-quasi-ordered} by the \kl{infix relation}.
\end{conjecture}

>>>>

The following result due to Kuske shows the structural complexity of the \kl{infix relation}

\begin{lemma}{\cite[Lemma 5.1]{DBLP:journals/ita/Kuske06}}
    \label{infix-embedding:thm}
    Let $(X, \preceq)$ be a \kl{partially ordered} set,
    and $\Sigma$ be an alphabet with at least two letters.
    Then the following are equivalent:
    \begin{enumerate}
        \item \label{infix-embedding-embeds:item} 
            $X$ \kl{embeds into} $(\Sigma^*, \infleq)$,
        \item \label{infix-embedding-count:item}
            $X$ is countable, and for every $x \in X$,
            its \kl{downwards closure}
            $\dwset[\preceq]{x}$ is finite.
    \end{enumerate}
\end{lemma}
\begin{figure}
    \centering
    \includestandalone[width=\linewidth]{fig/infix-encoding-standalone}
    \caption{Representation of the \kl{subword relation} for $\set{a,b}^*$
        inside the \kl{infix relation} for $\set{a,b,\#}^*$
        using the encoding of \cref{infix-embedding:thm}, restricted to words
        of length at most $3$. To obtain smaller words,
        we replaced $D_i$ by $\max D_i$ in this construction.
    }
    \label{infix-embedding:fig}
\end{figure}

@@

In the upcoming proof of \cref{infix-amalgamation:thm}, we will use the notion
of \intro{bounded language} for languages $L \subseteq \Sigma^*$ such that
there exists words $w_1, \dots, w_n$ satisfying $L \subseteq w_1^* \cdots
w_n^*$. This will be useful for us to derive the inclusion of a language $L$
into a union of \kl{chains}. Note that we isolate this intermediate result in a
separate theorem that could be applied beyond \kl{amalgamation systems}.

\begin{theorem}
    \label{bounded-language:thm}
    Let $L$ be a \kl{bounded language} of $\Sigma^*$. Then,
    $L$ is a \kl{well-quasi-order} when endowed with the 
    \kl{infix relation} if and only if it is a finite union of \kl{chains}.
\end{theorem}
\begin{proof}
    Let $w_1, \dots, w_n$ be such that
    $L \subseteq w_1^* \cdots w_n^*$.
    Let us define $m \defined \max \setof{\card{w_i}}{1 \leq i \leq n}$

    Now, admissible languages are ready to be pumped according to
    \cref{pumping-periods:lem}. For every admissible language,
    the size of a word that is not iterated is at most
    $N_0 \times m$ by definition, and we conclude that:
    \begin{equation}
    	\label{amalgamation-infix-wqo-inclusion:eqn}
        L \subseteq 
        \bigcup_{x,y \in \Sigma^{\leq n}}
        \bigcup_{u \in \Sigma^{\leq m \times N_0}}
        \InfPeriodChain{x} u \InfPeriodChain{y}
        \cup
        \InfPeriodChain{x} u
        \cup
        u \InfPeriodChain{x}
        \quad .
    \end{equation}
\end{proof}

@@

For the next result, we impose a mild restriction by requiring the language class to be closed under rational transductions. This property holds for most common language classes, including all known \kl{amalgamation systems}.

\begin{theorem}
	\label{infix-wqo-is-emptiness:thm}
	Let $\mathcal{C}$ be a class of languages recognized by amalgamation systems and closed under rational transductions. Then the following are equivalent:

	\begin{enumerate}
		\item\label{wqo-infix-decidable} \kl[well-quasi-order]{Well-quasi-orderedness} of the \kl{infix relation} is decidable.
		\item\label{emptiness-decidable} Emptiness is decidable.
	\end{enumerate}
\end{theorem}

\begin{proof}
	\cref{emptiness-decidable} $\Rightarrow$ \cref{wqo-infix-decidable}. We aim to make the inclusion test of \cref{amalgamation-infix-wqo-inclusion:eqn} effective.

	Let $R = \bigcup_{x,y \in \Sigma^{\leq n}} \bigcup_{u \in \Sigma^{\leq m \times N_0}} \InfPeriodChain{x} u \InfPeriodChain{y} \cup \InfPeriodChain{x}u \cup u\InfPeriodChain{x}$. For any concrete values of the bounds $n$, $m$, and $N_0$, this language is regular. As $\mathcal{C}$ is closed under rational transductions, we can therefore reduce the inclusion to emptiness of $L \cap \overline{R}$. However, we need to find these bounds first.

	To determine values for $n$ and $m$, we first test if $L$ is bounded. As the algorithm in \cite[Section 4.2]{ASZZ24} is effective, this yields words $w_1, \ldots w_n$ such that $L \subseteq w_1^* \cdots w_n^*$ and therefore yields also the bounds in questions.

	To determine the value for $N_0$, we first computer maximal subsets $S \subseteq [n]$ such that $w_i$ for $i \in S$ are simultaneously unbounded. We do this by applying a transduction mapping $w_i$ to some fresh letter $a_i$ if $i \in S$ and $\varepsilon$ otherwise, and testing for simultaneous unboundedness \cite[Section 4.1]{ASZZ24}. Given a candidate bound $n_0$ for $N_0$, we construct the languages $L_{S,n_0} = w_1^{\circ_1}\cdots w_n^{\circ_n}$ where $\circ_i = *$ if $i \in S$ and $\circ_i = \leq n_0$ otherwise. We then test if $L \subseteq \bigcup_{S \text{ maximal}} L_{S,n_0}$. If yes, $n_0$ is our value for $N_0$, otherwise, we increase it and repeat the construction. As we chose maximal sets $S$, this procedure will eventually terminate.

	\cref{wqo-infix-decidable} $\Rightarrow$ \cref{emptiness-decidable}. We 
	consider the transduction $\mathcal T = \Sigma^* \times \{a, b\}$. Then 
	for any language $L \in \mathcal C$, $\mathcal TL$ is well-quasi-ordered 
	by infix if and only if $L$ is empty.
\end{proof}
